---
title: "Reproducible Research: Project 1"
author: "Duncan Munslow"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
```
  
#### Libraries Used
```{r libraries, message=FALSE}
library(ggplot2)
library(dplyr)
```

#### 1. Code for reading in dataset/processing data
```{r read_process, cache = TRUE}

# Save file URL
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"

# Download and extract data into working directory if it does not already exist
if(!file.exists("proj1.zip")){
download.file(fileURL, "./proj1.zip")
unzip("./proj1.zip", exdir = "./proj1data")
}

# Read data into R
activityData <- read.csv("./proj1data/activity.csv", header =T)

# Convert date variable to correct format
activityData$date <- as.Date(as.character(activityData$date, "%Y-%m-%d"))
```


#### 2. Histogram of the total number of steps taken each day
```{r steps_per_day_hist}

# Group data by date(day), and then calculate steps per day - ignoring missing values.  
# Store results in stepsPerDay data frame
activDataGroup <- group_by(activityData, date)
stepsPerDay <- summarize(activDataGroup, stepsPerDay = sum(steps, na.rm = T))
                            
# Create histogram with 15 bins
qplot(stepsPerDay$stepsPerDay, bins = 15)

```

#### 3. Mean Steps per day
```{r mean_median_steps}

mean(stepsPerDay$stepsPerDay)
median(stepsPerDay$stepsPerDay)

```

We note that the mean steps per day is 9354.23, and the median steps is 10395. 

#### 4. Time Series plot of average number of steps taken
```{r daily_activity_plot}

# Group the data by interval
byInterval <- group_by(activityData, interval)

# Calculate the average steps per interval - ignoring missing values
dailyActivity <- summarize(byInterval, meanSteps = mean(steps, na.rm = T))

# Plot time series with ggplot 2, using the dailyActivity data
qplot(interval, meanSteps, data = dailyActivity, geom ='line' )

```

#### 5. 5-minute Interval with maximum steps
```{r max_steps}
dailyActivity[which(dailyActivity$meanSteps == max(dailyActivity$meanSteps)),]
```
We note that the interval with the maximum (on average) steps is 835, with 206.2 steps 

#### 6. Code to describe and show a strategy for imputing missing values
```{r count_missing_values}

sum(is.na(activityData$steps))

```

We note that there are 2304 missing values in the data set.  My strategy to impute
missing values is to replace "NA" values with the average steps for that interval
```{r impute_missing_values}

# create new data set with imputed values to preserve original data set
activeDataImputed <- activityData

# Replace missing values with interval average for that period using matrix operations
activeDataImputed$steps[is.na(activeDataImputed$steps)] <- 
    dailyActivity$meanSteps[match(activeDataImputed$interval,dailyActivity$interval)]

sum(is.na(activeDataImputed$steps))

```

We see that the imputation process has removed all missing values.

#### 7. Histogram of steps per day with imputed values
```{r histogram_imputed}

imputedGrouped <- group_by(activeDataImputed, date)

imputedStepPerDay <- summarize(imputedGrouped, steps_per_day = sum(steps))

# Make histogram with 15 bins as in orginal plot
qplot(imputedStepPerDay$steps_per_day, bins = 15)
```

We see that the number of days with 0 steps has decreased drastically  

##### Mean and Median imputed steps
```{r mean_median_steps_imputed}
mean(imputedStepPerDay$steps_per_day)
median(imputedStepPerDay$steps_per_day)

```
Interestingly, the mean and median steps values are exactly equal to one another with 10766.19 steps  


#### 8. Panel plot - Average # of steps Weekday vs. Weekend
```{r weekday_coding}

# Create new data frame with imputed data to preserve orginal data
patternData <- activeDataImputed

# create new variable "daynames" using weekdays function, which assigns the day to each interval
patternData$daynames <- weekdays(patternData$date)

# Create charactor vector with weekend names 
weekends <- c("Saturday", "Sunday")

# Create indicator for weekend and weekday variable types
patternData <- mutate(patternData, 
                      dayType = ifelse(daynames %in% weekends, 
                                       "weekend", "weekday" ))

# Group the data by dayType and interval, and calculate the average steps by 
# day type and interval
patternGrouped <- group_by(patternData, dayType, interval)
patternSummarized <- summarize(patternGrouped, meanSteps = mean(steps))

```

#### Plot 
```{r weekend_vs_weekday_plot}

# create ggplot object using the summarized data
gg <- ggplot(data = patternSummarized, aes(x = interval, y = meanSteps, color = dayType))

# add facets for weekend and weeday and the lines for each to the gg object
gg <- gg + facet_grid(dayType~.) + geom_line()

# Print object
gg

```